# Assignment 6 - Analysis of longitudinal data


```{r}
date()
```
## 0.

Reflections: doing this the last day. Has been exhausting; been sick with a tenacious flu, and the baby doesn't like to sleep. Reflecting more on the whole course than just this week. I had used R briefly earlier, but for very different uses, and in a very different way. And of course some 6 years ago or so. So everything was not new, yet new. I found out that I need to grasp statistical concepts more firmly - my understanding in them is on too shaky a ground to make me feel comfortable doing analyses. 

Overall a good course in such sense that it gave me glimpses of something else. Yet, I cannot say that I would master these methods after this brief intro. So, perhaps next I need to take deeper looks into statistical analyses on the ground level and R in connection to them. My approach seems to be inverse: starting from the more complicated, advancing to the simpler stuff. Yet, I think that the analyses I aim to do with R might benefit from this course. I just need some more time with the fundamentals. Perhaps more hands-on "DIY"-thing would have been better: this course was intro, and as such, entailed lots of copying of ready code. That of course is what we all do, but to get something to stick to ones spine, a workflow to become the norm, one has to really hack it deep into the ape-brain. And for me the only way to achieve this is by repetition. So, perhaps the next step for me is doing some reps to get some gains on them old muscles; here in this course I had the PT to give me pointers on the technique.

Onwards to the assignment. This seems very applicative. But a lot of work, and I have only a few hours. I'll do what I can; first the graphs and such, and if I have time I will add the verbal analysis.

### EDIT:
Just couldn't keep my eyes open any longer, so did not finish the assignment properly. Most of the charts are done, but what is lacking is the analysis. Also the neat and tidy formulations to R Markdown are not done: some of the explanations are inside code blocks.

**Apologies in advance**, I'm afraid this will not be that easy to follow or give points.




Here I'll load all the necessary libraries
```{r}
library(tidyverse)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lme4)
```



# 1. Analyses of Chapter 8 of MABS-book, but with RATS -data.

Import data from the data wrangling set, check everything is ok.

```{r}
RATSL <- read_csv("data/rats.csv")

glimpse(RATSL)

# as evident from the glimpse, gotta redo the categorical variables into factors

RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)


```
Okey dokey. The assignment was to reproduce the charts and tables of MABS-book chapter 8 with RATS-data. I will go through the charts and tables by their numbers in the book, and redo them if possible. Analysis will be added as the final stage, if I have time.

NB! There are some charts / tables that I will not reproduce, since they are not relevant for the analysis. Such tables are for instance an example of the first 40 rows of data, and Table 8.2 "Possible summary measures", which is not calculated from the data.



IF NEED BE My data *RATSL* is in long form, which might not be the most useful for all the things needed. So, just in case, I will also reload the raw data in wide form. 

```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", 
                   header = TRUE, sep = '\t')
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
str(RATS)
```





## TABLE 8.1 BPRS Measurements from 40 Subjects 

-> this is just giving overview of the data, we're not going to repeat it.



## FIGURE 8.1 Individual response profiles by treatment group for the BPRS data.

```{r}

# has to have the linetype = ID for this graph to work. 

ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) + 
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
 
```

## FIGURE 8.2
Individual response profiles for BPRS data after standardization.

```{r}
# same as above, but with standardized values

RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) + 
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "standardized weight")
 



```

## FIGURE 8.3
Mean response profiles for the two treatment groups in the BPRS data.

```{r}
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight) ) %>%
  ungroup()

# Glimpse the data
# glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

## FIGURE 8.4
Boxplots for the BPRS data.

```{r}

glimpse(RATSL)

ggplot(RATSL, aes(x = WD, fill = Group, y = Weight))+ 
  geom_boxplot()



```



## TABLE 8.2
Possible Summary Measures] --> This is something else, not needed



## FIGURE 8.5
Boxplots of mean summary measures for the two treatment groups in the
BPRS data.

```{r}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0)
RATSL_1 <- RATSL %>%
  group_by(Group, ID) %>%
  summarise(mean = mean(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL_1)

# Draw a boxplot of the mean versus treatment
ggplot(RATSL_1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "blue") +
  scale_y_continuous(name = "mean(weight)")


```


## FIGURE 8.6
Boxplots of mean summary measures for the two treatment groups in the
BPRS data, without the outlier shown in Figure 8.5.

```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL_2 <- RATSL %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  filter(mean < 590) %>%
  ungroup()

  
ggplot(RATSL_2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "blue") +
  scale_y_continuous(name = "mean(Weight)")
```

I removed just one outlier; not sure if here was intended to remove all of them.


All of the three following are "grouped together" for the Anova -analysis

## TABLE 8.3
Results From an Independent Samples t-test on the Mean Summary
Measure for the BPRS Data, Without the Outlier Shown in Figure 8.5

## TABLE 8.4
Results from an Analysis of Covariance of the BPRS Data with Baseline
BPRS and Treatment Group as Covariates

## TABLE 8.6
Results from an Independent Samples t-test for the Mean Summary Measure
Used on the Data Partially Shown in Table 8.5. (a) Leaving Out Subjects
with Any Missing Value, (b) Mean of Available Values for Each Subject

```{r}
# Not sure what needs to be done here, so I'll recreate what was in the Exercise

# t.test(mean ~ Group, data = RATSL11, var.equal = TRUE) NOT suitable, because three factors, not two: just Anova, then

# Add the baseline from the original data as a new variable to the summary data
RATSL_3 <- RATSL_1 %>%
  mutate(baseline = RATS$WD1) # The original raw RATS data was needed after all!

RATSL_3 <- RATSL_3 %>%
  filter(mean < 590)
  

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL_3)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```


# 2. Analyses of Chapter 9 of MABS-book, but with BPRS -data.


Load data from data wrangling

```{r}
BPRSL <- read_csv("data/bprs.csv")

glimpse(BPRSL)

# Same as above, gotta redo the categorical variables into factors

BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

```

Same thing here as above, just in case loading the raw data 

```{r}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", 
                   sep =" ", header = T)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
str(BPRS)

```

Same thing as above, will do the charts and tables (and list them here with reference to the book), but did not have time to write analysis.

**NOTE:** all the chart "names" will refer to Rats, even though I am using BPRS here, because the chart names came from the book. 


##TABLE 9.1
Body Weights of Rats Recorded Over a 9-Week Period


This is just viewing the data, i.e.
```{r}
view(BPRS)
```



## TABLE 9.2
Long Form of the Data for the First Two Rats in Group 1 in Table 9.1

```{r}

#already done in wrangling, just glimpsing here

glimpse(BPRSL)
```


## FIGURE 9.1
Plot of weight against time for rat data, ignoring the repeated-measures structure
of the data but identifying the group to which each observation belongs.


RRrright... With the BPRS -data this is probably close enough 

```{r}
ggplot(BPRSL, aes(x=weeks, y = bprs, group = treatment, colour = treatment)) + 
  geom_point()
```



## TABLE 9.3
Results from Fitting a Linear Regression Model to Rat Data with Weight as
Response Variable, and Group and Time as Explanatory Variables, and
Ignoring the Repeated-Measures Structure of the Data

```{r}
BPRSL_week <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))


BPRSL_lm <- lm(bprs ~ week + treatment, data = BPRSL_week)
summary(BPRSL_lm)
```



## FIGURE 9.2
Plot of individual rat growth profiles.

```{r}

ggplot(BPRSL_week, aes(x = week, y = bprs, linetype = subject, colour = subject)) +
  geom_line() + 
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "bprs")
```




## FIGURE 9.3
Scatterplot matrix of repeated measures in rat growth data.


```{r}
BPRSL_week_only <- BPRSL_week[, -3]
pairs(BPRSL_week_only)
```



## TABLE 9.4
Results from Fitting Random Intercept Model, with Time and Group as
Explanatory Variables, to Rat Growth Data

```{r}
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL_week_only, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```



## TABLE 9.5
Results from Fitting the Random Intercept and Slope Model, with Time
and Group as Explanatory Variables, to Rat Growth Data

```{r}
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL_week_only, REML = FALSE)
summary(BPRS_ref1)
```
```{r}
anova(BPRS_ref1, BPRS_ref)
```



## TABLE 9.6
Results from Fitting the Random Intercept and Slope Model that Allows for
a Group × Time Interaction to Rat Growth Data

```{r}
BPRS_ref2 <-lmer(bprs ~ week * treatment + (week | subject), data = BPRSL_week_only, REML = FALSE)
summary(BPRS_ref2)
```



## FIGURE 9.4
Fitted growth rate profiles from the interaction model and observed growth
rate profiles.

```{r}
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to RATSL
BPRSL_f <- BPRSL_week_only %>% mutate("Fitted" = Fitted)

# draw the plot of RATSL with the Fitted values of weight
ggplot(BPRSL_f, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  facet_grid(. ~ treatment, labeller = label_both) +
  #scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "none")

# must be divided into two, otherwise will look jagged


```

## TABLE 9.7
First Five Patients in Each Treatment Group of the “Beat the Blues” (BtB)
Clinical Trial of CBT for Depression

--> not interesting, not done


## FIGURE 9.5
Box plots of BDI scores by occasion of recording and treatment group.

```{r}

# this is just the same exercise as in Exercise 6 copied beneath, we have pretty much done every meaningful chart


BPRSL8S <- BPRSL_week %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()


# Draw a boxplot of the mean versus treatment
library(ggplot2)
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")


```



## FIGURE 9.6
Scatterplot matrix of BDI scores.

Emm... already done.


## TABLE 9.8
Results from Fitting a Multiple Linear Regression Model to BtB Data
Assuming the Repeated Measurements of BDI are Independent

Same, repeating the same from above (the book changed datasets in the middle)


## The END

Thank you for reading thus far, and apologies for the really awful format of this last assignment. Just didn't have time, due to life. Score accordingly.  Peace out!
